Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-02 mr-review.Rmd
updated ms

Great interest has developed in inferring causal relationships between complex 
traits in observational human genetics studies. Throughout the manuscript, we 
use the term "complex trait" to refer to a trait that doesn't appear to be 
inherited in a Mendelian fashion. Typically, such complex traits such as gene expression levels
are inherited in a polygenic manner, with many genes influencing the trait.

Genome-wide association studies (GWAS) present unique opportunities for 
discovery of causal relationships via mendelian randomization. A genome-wide association study 
interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait. A 
SNP is a genetic marker defined by nucleotide base identity at a single genomic position.

Methods for Mendelian randomization have appeared at a rapid and accelerating pace since the 
appearance of the first GWAS.

@@

@@

There are several ways that real data may violate the three MR assumptions. These include ___, ____, and ___.

@@

>>>>

Great interest has developed in inferring causal relationships between complex 
traits in observational human genetics studies. Discovery of such relationships is crucial to 
enhancing our understanding of the biology of health and disease. Two-sample 
Mendelian randomization (MR) studies are a form of instrumental variable analysis in which
genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments for 
inferring causal effects of an exposure variable on an outcome variable. They are termed 
"two-sample" MR studies because the exposure variable and the outcome variable are measured on 
two non-overlapping cohorts of subjects. Importantly, the SNPs are measured in both sets of subjects.

Throughout the manuscript, we 
use the term "complex trait" to refer to a trait that doesn't appear to be 
inherited in a Mendelian fashion. Typically, such complex traits, for example, gene expression levels,
are inherited in a polygenic manner, with many genes influencing the trait.

Genome-wide association studies (GWAS) present unique opportunities for 
discovery of causal relationships via MR. A GWAS
interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait. A 
SNP is a genetic marker defined by nucleotide base identity at a single genomic position.

```{r, article-counts, fig.cap= "Upward trend in article counts by year for keyword searches: 1. mendelian randomization and 2. genome-wide association study."}
knitr::include_graphics("figures/mr-gwas-counts-scatter.png")
```

Methods for Mendelian randomization have appeared at a rapid and accelerating pace since the 
appearance of the first GWAS (Figure \@ref(fig:article-counts)).

@@

~~Investigating causal relationship among complex traits and identifying causal risk
factors are an important first step towards understanding the biology of diseases. A
common statistical tool for performing such causal inference in observational studies is
Mendelian randomization (MR). MR is a form of instrumental variable analysis that
uses SNPs to serve as instruments for inferring the causal effect of an exposure variable
on an outcome variable(1). MR requires only summary statistics from genome-wide
association studies (GWASs) and is often performed in a two-sample study setting
where the exposure variable and the outcome variable are measured in two separate
studies(2). With the abundant availability of GWAS summary statistics, numerous MR
analyses are being carried out, identifying important causal risk factors for various
common diseases. These MR studies are facilitated by many recently developed MR
methods that include the inverse variance weighted (IVW) method, MR-Egger(3),
median-based regression(4), BWMR(5), RAPS(6), MRMix(7), CAUSE(8), to name a
few. Different MR methods differ in their modeling assumptions and inference
algorithms, but the majority of them encounter two important modeling and algorithmic
challenges that have so far limited the effectiveness of MR analysis.
First, almost all existing MR methods rely on a pre-selected set of independent
SNPs to serve as instruments for MR analysis. The instruments are selected to be
independent from each other to ensure the validity of the statistical inference framework
used in many common MR methods such as IVW. The independent SNPs are often
selected through linkage disequilibrium (LD) clumping, a procedure that first ranks
SNPs based on their marginal association evidence with the exposure variable and then
retains SNPs that are not in high LD with the SNPs on top of the ranking list. Using LD
clumping to select SNPs may be suboptimal, however, as the selected SNPs may only
epresent tagging SNPs that are in LD with the causal SNPs rather than the causal ones
themselves. Using tagging SNPs instead of the causal ones to serve as instruments can
reduce the power of MR analysis. In addition, perhaps more importantly, selecting
independent SNPs for MR analysis may not be ideal either, as complex traits can be
influenced by multiple causal SNPs residing in the same local region that are in
potential LD with each other. Consequently, selecting independent SNPs may only
capture a small proportion of the phenotypic variance in the exposure variable, again
leading to a loss of power in the subsequent MR analysis (1, 2, 9, 10). Indeed, in the
parallel research field of transcriptome-wide association studies, it has been well
documented that incorporating correlated SNPs can substantially improve analysis
power than using independent SNPs only(11-14). Therefore, incorporating correlated
SNPs and developing effective approaches to select instruments among them are
important to fully captivate the potential of MR.
Second, only a limited number of MR methods model horizontal pleiotropy and
even fewer can effectively control for it during MR analysis(15). Horizontal pleiotropy
occurs when the SNP instruments exhibit effects on the outcome through pathways
other than the exposure. Horizontal pleiotropy has been widely observed in complex
trait analysis(13, 15) and often comes in two distinct types. The first type of horizontal
pleiotropy arises through paths independent of the exposure, with the resulting
horizontal pleotropic effects being independent of the SNP effects on the exposure. The
second type of horizontal pleiotropy arises through unobserved exposure-outcome
confounders and induces correlation between the horizontal pleotropic effects and the
SNP effects on the exposure. The presence of either type of horizontal pleiotropy
violates standard MR modeling assumptions and can lead to biased causal effect
estimates and increased false discoveries. Early MR analyses control for horizontal~~

@@

@@

There are several ways that real data may violate the three MR assumptions. These include .

## Sensitivity analysis for MR with one instrument and one outcome

Sensitivity analysis is recommended and widely used in MR studies because of the inability to verify 
the three MR assumptions with observational data. The goal of a sensitivity analysis is to gain insight 
into how the results might differ were we able to verify the MR assumptions.
